---
title: "BIOL 812 Final Project"
output:
  word_document: default
  html_document: default
date: "2025-04-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup
```{r}
library(ggplot2) # plotting library
library(dplyr) # data management
library(lme4) # linear mixed-effects models using Eigen and S4
library(lmtest) # Lagrange multiplier test
library(stringr) 
library(tidyr)
library(RColorBrewer) # colour graphics
library(sjPlot) # data visualization
source("http://bit.ly/theme_pub") # set custom plotting theme
theme_set(theme_pub())
```

# Download Data
```{r}
GarDat = read.csv("CommonGarden_Data_With_ENV (2).csv")
```

# Set random effect (Block) as factor
```{r}
GarDat <- GarDat %>% 
  mutate(Block=as.factor(Block))
```

# Linear mixed effects models with fixed environmental effects (Module 3)
```{r}
GrowthMod1 <-  lmer(Height_Growth_Rate ~ Latitude + Precipitation_of_Wettest_Month + Mean_Annual_Temp + Temperature_Seasolity + (1 | Block), na.action = na.exclude, data = GarDat)
GrowthMod2 <- lmer(Height_Growth_Rate ~ Precipitation_of_Wettest_Month + Mean_Annual_Temp + Temperature_Seasolity + (1 | Block), na.action = na.exclude, data = GarDat)
GrowthMod3 <-lmer(Height_Growth_Rate ~ Latitude + Mean_Annual_Temp + Temperature_Seasolity + (1 | Block), na.action = na.exclude, data = GarDat)
GrowthMod4 <- lmer(Height_Growth_Rate ~ Latitude + Precipitation_of_Wettest_Month + Temperature_Seasolity + (1 | Block), na.action = na.exclude, data = GarDat)
GrowthMod5 <- lmer(Height_Growth_Rate ~ Latitude + Precipitation_of_Wettest_Month + Mean_Annual_Temp + (1 | Block), na.action = na.exclude, data = GarDat)
GrowthMod6 <- lmer(Height_Growth_Rate ~ Mean_Annual_Temp + Temperature_Seasolity + (1 | Block), na.action = na.exclude, data = GarDat)
GrowthMod7 <- lmer(Height_Growth_Rate ~ Latitude + Precipitation_of_Wettest_Month + (1 | Block), na.action = na.exclude, data = GarDat)
GrowthMod8 <- lmer(Height_Growth_Rate ~ Latitude + Temperature_Seasolity + (1 | Block), na.action = na.exclude, data = GarDat)
GrowthMod9 <- lmer(Height_Growth_Rate ~ Latitude + (1 | Block), na.action = na.exclude, data = GarDat)
GrowthMod10 <- lmer(Height_Growth_Rate ~  Precipitation_of_Wettest_Month + (1 | Block), na.action = na.exclude, data = GarDat)
GrowthMod11 <- lmer(Height_Growth_Rate ~ Mean_Annual_Temp + (1 | Block), na.action = na.exclude, data = GarDat)
GrowthMod12 <- lmer(Height_Growth_Rate ~ Temperature_Seasolity + (1 | Block), na.action = na.exclude, data = GarDat)

lrtest(GrowthMod1, GrowthMod2, GrowthMod3, GrowthMod4, GrowthMod5, GrowthMod6, GrowthMod7, GrowthMod8, GrowthMod9, GrowthMod10, GrowthMod11, GrowthMod12)
```
# Likelihood ratio test demonstrates statistical significance for all models but GrowthMod1

# AIC and BIC for model selection - GrowthMod7
```{r}
AIC(GrowthMod1, GrowthMod2, GrowthMod3, GrowthMod4, GrowthMod5, GrowthMod6, GrowthMod7, GrowthMod8, GrowthMod9, GrowthMod10, GrowthMod11, GrowthMod12)
BIC(GrowthMod1, GrowthMod2, GrowthMod3, GrowthMod4, GrowthMod5, GrowthMod6, GrowthMod7, GrowthMod8, GrowthMod9, GrowthMod10, GrowthMod11, GrowthMod12)
```
# Model selection for height growth rate identifies GrowthMod7 to be best fit

# Plotting Best Model for Height Growth Rate
```{r}
p <- sjPlot::plot_model(GrowthMod7, title = "Effect of Fixed Effects on Height Growth Rate")
p+theme_sjplot()

tab_model(GrowthMod7,
          show.re.var = TRUE, dv.labels = "Effect of Fixed Effects on Height Growth Rate")
```
# Plot presents a negative relationship between latitude and Height Growth rate (Figure of best model as per AIC)


# Plotting Model with all variables (model not significant or the best as per AIC)
```{r}
p2 <- sjPlot::plot_model(GrowthMod1, title = "Effect of Fixed Effects on Height Growth Rate")
p2+theme_sjplot()

tab_model(GrowthMod1,
          show.re.var = TRUE, dv.labels = "Effect of Fixed Effects on Height Growth Rate")
```
# Plot of all variables demonstrating a negative relationship between fixed variables and Height Growth rate

# ANOVA on most significant variables from LMM Height Growth Rate (Module 2)
```{r}
summary(GrowthMod7)
anova(GrowthMod7)
```

# Evaluating how most significant LMM variables affect Germination using ttests (Module 1)
```{r}
t.test(GarDat$Latitude~GarDat$Germited)
t.test(GarDat$Precipitation_of_Wettest_Month~GarDat$Germited)
```
# Two sample t-tests demonstrate statistical significance between both variables and Germination

# Plotting germination data
```{r}
ggplot(GarDat, aes(x=as.factor(GarDat$Germited), y=GarDat$Latitude, color=as.factor(GarDat$Germited))) +
  geom_boxplot(alpha=0.3) + labs(title="Germination with respect to latitude", x="Germination", y="Latitude")

ggplot(GarDat, aes(x=as.factor(GarDat$Germited), y=GarDat$Precipitation_of_Wettest_Month, color=as.factor(GarDat$Germited))) +
  geom_boxplot(alpha=0.3) + labs(title="Germination with respect to precipitation", x="Germination", y="Precipitation of Wettest Month")
```

